The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-18 manuscript.Rmd
fix author order and add @daroczig to contribution section
some housekeeping

For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g. the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

@@

**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g. [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).

@@

There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are needed, e.g. if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.

@@

_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).

@@

```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage an instances in the cloud",
  "interact", "Interact with an instance (e.g. file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

@@

Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
Some packages, e.g. \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.

@@

Second, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
That can be a challenge for some users.
The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g. using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly seperates test environment, test code, and current working copy of the code.

@@

Nonetheless, at least on the level of R packages some **consolidation** seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

New features, which make complex workflows accessible and reproducible, and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.

>>>>

For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

@@

**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).

@@

There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.

@@

_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).

@@

```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage an instances in the cloud",
  "interact", "Interact with an instance (e.g., file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

@@

Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.

@@

Second, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
That can be a challenge for some users.
The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly seperates test environment, test code, and current working copy of the code.

@@

Nonetheless, at least on the level of R packages some **consolidation** seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g., `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

New features, which make complex workflows accessible and reproducible, and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.
